---
title: "Processmine — Process Analysis Walkthrough"
subtitle: "Discovery · Conformance · Performance · Variants"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 4
)
```

## Setup

Load the processmine R package. If you have it installed, replace the `load_all` call with `library(processmine)`.

```{r load}
suppressPackageStartupMessages({
  devtools::load_all("../R", quiet = TRUE)
})
```

---

## 1 · Read and validate the event log

The sample log is a synthetic order-to-cash process: 15 cases covering three
path variants — happy path (no credit check), credit-checked approvals, and
rejections.

```{r read-xes}
log <- read_xes("../data/sample_logs/bpi2012_sample.xes")

cat("Cases  :", length(unique(log$case_id)), "\n")
cat("Events :", nrow(log), "\n")
cat("Period :", format(min(log$timestamp)), "→", format(max(log$timestamp)), "\n")
```

```{r log-head}
head(log[, c("case_id", "activity", "timestamp", "resource", "lifecycle")])
```

```{r activity-counts}
sort(table(log$activity), decreasing = TRUE)
```

---

## 2 · Process Discovery

### 2.1 Directly-Follows Graph

The DFG counts how many times activity B immediately follows activity A within
the same case. `freq` is relative to total events.

```{r dfg}
dfg <- discover_dfg(log)

cat("Activities      :", paste(dfg$activities, collapse = ", "), "\n")
cat("Start activities:", paste(dfg$start_activities, collapse = ", "), "\n")
cat("End activities  :", paste(dfg$end_activities,   collapse = ", "), "\n\n")

dfg$edges
```

Visualise the edge weights with a simple dot plot.

```{r dfg-plot}
edges <- dfg$edges
edges$label <- paste0(edges$from, "\n→ ", edges$to)

par(mar = c(5, 10, 2, 2))
barplot(
  rev(edges$n),
  names.arg = rev(edges$label),
  horiz     = TRUE,
  las       = 1,
  col       = "#4e79a7",
  xlab      = "Edge count",
  main      = "DFG — directly-follows edge counts"
)
```

### 2.2 Heuristics Net

The heuristics miner retains only edges whose dependency measure
`dep(A,B) = (|A→B| − |B→A|) / (|A→B| + |B→A| + 1)` meets a threshold.
Low-frequency noise edges are filtered out.

```{r heuristics}
hnet <- discover_heuristics(log, dependency_threshold = 0.8)

cat("Edges retained:", nrow(hnet$edges), "of", nrow(dfg$edges), "\n\n")
hnet$edges[, c("from", "to", "n", "dependency")]
```

---

## 3 · Conformance Checking

Token replay measures how well each case follows the discovered model.
A fitness of `1.0` means every transition in the case exists in the model.

We build the reference model from the happy-path cases (OTC-001 to OTC-010)
and then replay all 15 cases against it.

```{r conformance}
ref_cases  <- log[log$case_id %in% paste0("OTC-0", sprintf("%02d", 1:10)), ]
ref_model  <- discover_dfg(ref_cases)
conf       <- conformance_tokenreplay(log, ref_model)

cat("Mean fitness  :", round(conf$summary$mean_fitness,   3), "\n")
cat("Median fitness:", round(conf$summary$median_fitness, 3), "\n")
cat("Perfect fit   :", conf$summary$n_fitting_cases, "of", conf$summary$n_cases, "cases\n")
```

```{r conformance-per-case}
conf$per_case
```

Cases OTC-011 to OTC-013 follow the happy path after a credit check, so they
have partial fitness (the Check Credit → Approve Order transition exists in the
log but not in the happy-path reference model). OTC-014 and OTC-015 are
rejected and have zero fitness against the happy-path model.

```{r conformance-plot}
pc <- conf$per_case
pc$colour <- ifelse(pc$fitness >= 1.0, "#59a14f",
             ifelse(pc$fitness >  0.0, "#f28e2b", "#e15759"))

par(mar = c(8, 4, 3, 1))
barplot(
  pc$fitness,
  names.arg = pc$case_id,
  col       = pc$colour,
  las       = 2,
  ylim      = c(0, 1.1),
  ylab      = "Fitness",
  main      = "Token replay fitness per case",
  border    = NA
)
abline(h = 1, lty = 2, col = "grey50")
legend("bottomright",
       fill   = c("#59a14f", "#f28e2b", "#e15759"),
       legend = c("Perfect fit", "Partial fit", "Zero fit"),
       bty    = "n")
```

---

## 4 · Performance Analysis

### 4.1 Throughput time

```{r throughput}
tp <- performance_throughput(log, unit = "hours")
tp
```

```{r throughput-summary}
cat("Mean throughput  :", round(mean(tp$throughput), 1), "hours\n")
cat("Median throughput:", round(median(tp$throughput), 1), "hours\n")
cat("Min / Max        :", round(min(tp$throughput), 1), "/",
    round(max(tp$throughput), 1), "hours\n")
```

```{r throughput-plot}
par(mar = c(8, 5, 3, 1))
barplot(
  tp$throughput,
  names.arg = tp$case_id,
  las       = 2,
  col       = "#76b7b2",
  ylab      = "Throughput (hours)",
  main      = "Case throughput time",
  border    = NA
)
```

### 4.2 Bottlenecks

Which transitions take the longest on average?

```{r bottlenecks}
bn <- performance_bottlenecks(log)
bn$mean_hours <- round(bn$mean_duration_s / 3600, 1)
bn[, c("from_activity", "to_activity", "mean_hours", "n")]
```

The longest gap is always `Issue Invoice → Receive Payment` — payment arrives
days after invoicing, which is expected business behaviour. Everything within a
single working day is operational throughput.

### 4.3 SLA compliance

Define a 7-day (168-hour) SLA for order completion.

```{r sla}
sla <- performance_sla(log, sla_spec = list(limit = 168, unit = "hours"))

cat("Within SLA:", sum(sla$within_sla), "of", nrow(sla), "cases\n\n")
sla[, c("case_id", "throughput", "within_sla")]
```

Rejected cases (OTC-014, OTC-015) are short — they close quickly — so they
trivially meet the SLA. The few cases that breach it are the longer-running
credit-check approvals.

---

## 5 · Variant Analysis

```{r variants}
v <- variants(log)
v
```

```{r variants-plot}
par(mar = c(3, 22, 3, 1))
barplot(
  rev(v$n),
  names.arg = rev(v$variant),
  horiz     = TRUE,
  las       = 1,
  col       = c("#59a14f", "#4e79a7", "#e15759"),
  xlab      = "Case count",
  main      = "Process variants",
  border    = NA
)
```

```{r rare}
rare <- rare_paths(log, min_support = 0.15)
cat("Rare variants (< 15% support):", nrow(rare), "\n\n")
rare
```

---

## 6 · Write to Parquet for Python ML

Export the validated log so the Python notebook can pick it up for anomaly
detection, drift monitoring, and next-activity prediction.

```{r write-parquet}
out <- "../data/sample_logs/bpi2012_sample.parquet"
write_eventlog_parquet(log, out)
cat("Written:", out, "\n")
```

---

## Session info

```{r session}
sessionInfo()
```
